---
title: "Michigan PFAS Data"
format:
  html:
    toc: true
    self-contained: true
    embed-resources: true
execute:
  echo: false  # Do not show code
  warning: false  # Suppress warnings
  message: false  # Suppress messages
  output: true  # Show output
---

```{r}
#| include: false
set.seed(1994)

# Loading necessary packages
library(tidyverse)
library(gt)
library(skimr)
library(tidycensus)
library(maps)

# Function for creating data dictionary
data_dictionary <- function(myData, descripts) {
tibble(Variable = colnames(myData),
       Type = map_chr(myData, .f = function(x){typeof(x)[1]}),
       Description = descripts) |> 
  gt()
}
```

```{r PFAS_packages}
#| eval: true
#| include: false

library(tidyverse)
library(naniar)
library(janitor)
library(sf)

# Create function to add geoid variable given longitude and latitude for county-level data
add_geoid <- function(my_data) {
  # Create sf object to help add geoid column
my_data_sf <- my_data |> 
st_as_sf(coords = c("longitude", "latitude"))

# Set coordinate reference system (CRS) to match counties data
counties <- st_read("GeoFiles//cb_2018_us_county_500k.shp")
st_crs(my_data_sf) <- st_crs(counties)

ret_data <- my_data_sf |> 
  st_join(counties) |> 
    mutate(longitude = st_coordinates(geometry)[, 1],
           latitude = st_coordinates(geometry)[, 2]) |> 
  st_drop_geometry() |> 
  as_tibble() |> 
  dplyr::select(GEOID, longitude, latitude) |> 
  dplyr::rename(geoid = GEOID) |> 
  distinct() |> 
  right_join(my_data)

return(ret_data)
}
```

```{r PFAS_surface_clean}
#| eval: true
#| include: false

# Importing data on PFAS levels in surface water in Michigan
analyte_info <- read_csv("PFAS_Surface_Water_Sampling.csv") |> 
              dplyr::select(-c(X:Latitude, GlobalID, OBJECTID,
                               ends_with("Flag")))

surface_water <- read_csv("PFAS_Surface_Water_Sampling.csv") |> 
  clean_names() |> 
  dplyr::select(-ends_with("_flag")) |> 
  dplyr::select(-c(cas307244_pf_hx_a:cas919005144_adona_rl)) |> 
  bind_cols(analyte_info)

# Importing data on surface water data dictionary
surface_water_dictionary <- read_csv("surface_water_data_dictionary.csv")

# Creating long version of surface water data
surface_water_long <- surface_water |> 
  dplyr::rename(object_id = objectid) |> 
  dplyr::select(longitude, latitude, object_id, everything(), 
                -c(x:y), -ends_with("Flag"), -global_id) |> 
  pivot_longer(cols = colnames(analyte_info), 
               names_to = "analyte", 
               values_to = "analyte_value") |> 
  dplyr::select(-matrix, -unit, -object_id) |> 
  add_geoid()

# Saving to external CSV
write_csv(surface_water_long, file = "pfas_surface_water_long.csv")
```

```{r PFAS_public_clean}
#| eval: true
#| include: false

# Importing shape file data for public water supply PFAS levels data
public_water_shape <- st_read("Public_Water_Supply_Sampling_Hexbins_and_Results/Public_Water_Supply_Sampling_Hexbins.shp") |> 
  mutate(centroid = st_centroid(geometry),
         longitude = st_coordinates(centroid)[, 1],
         latitude = st_coordinates(centroid)[, 2]) |> 
  dplyr::select(HexID, longitude, latitude) |> 
  st_drop_geometry() |> 
  as_tibble()

# Importing data on public water supply PFAS levels
public_water <- read_csv("Public_Water_Supply_Sampling_Hexbins_and_Results.csv")

# Adding geoid, longitude, and latitude information to public water data
# and cleaning variable names
public_water_wide <- public_water |> 
  left_join(public_water_shape, by = "HexID") |> 
  dplyr::select(-ends_with(c("Result", "Flags"))) |> 
  clean_names() |> 
  dplyr::rename(object_id = objectid) |> 
  dplyr::select(object_id:sys_loc_code, longitude, latitude, everything()) |> 
  bind_cols(dplyr::select(public_water, ends_with("Result"))) |> 
  add_geoid()

# Pivoting public water data from wide to long format
public_water_long <- public_water_wide |> 
  pivot_longer(cols = c(ends_with(c("Result"))), 
               names_to = "analyte", 
               values_to = "analyte_value") |> 
  mutate(analyte = gsub(analyte, pattern = "Result", replacement = "")) |> 
  dplyr::select(-c(hex_id, 
                   wssn, loc_name:sys_loc_code, 
                   phase_code:task_type,
                   treatment_status:sys_sample_code,
                   lab_number:lab_sdg, 
                   object_id, position_source,
                   analytical_method,
                   sampling_results_count)) |> 
  mutate(system_type = fct_recode(system_type,
                               "Non-Community Water Supply (Adult Foster Care Provider)" = "ADFSTC",
                               "Non-Community Water Supply (Children's Camp)" = "CHLCMP",
                               "Non-Community Water Supply (Child Care Provider)" = "DAYCARE",
                               "Non-Community Water Supply (Industry)" = "INDUS",
                               "Non-Community Water Supply (Medical Care Provider)" = "MEDCAR",
                               "Non-Community Water Supply (Hotel or Motel)" = "MOTEL",
                               "Community Water Supply (for example Municipal Supply, Apartment, Nursing Home, Prison, etc.)" = "MUN",
                               "Office Building" = "OFFICE",
                               "Park" = "PARK",
                               "Residential" = "RESD",
                               "School" = "SCH",
                               "Tribal Lands" = "TRB"))


# Saving to external CSV
write_csv(public_water_long, file = "pfas_public_water_long.csv")
```


```{r PFAS_sites_import}
#| eval: true
#| include: false

pfas_site_path <- "pfas_sites.csv"

if(file.exists(pfas_site_path) == FALSE) {
# Importing data on identified sites from https://www.michigan.gov/pfasresponse
# Downloaded on 02/25/2025
pfas_sites <- read_csv("Michigan_PFAS_Sites.csv") |> 
  janitor::clean_names() |> 
  dplyr::select(facility:site_type, location:military, facility_date, site_background:anticipated_activities, -tier, -site_type) |> 
  add_geoid()

# Saving cleaned data
write_csv(pfas_sites, file = pfas_site_path)
} else {
  # Importing
  pfas_sites <- read_csv(pfas_site_path)
}
```

## Description

From the state of Michigan's website:

*PFAS are a large group of man-made chemicals that include perfluorooctanoic acid (PFOA) and perfluorooctanesulfonic acid (PFOS). PFAS have been used globally during the past century in manufacturing, firefighting and thousands of common household and other consumer products. These chemicals are persistent in the environment and in the human body – meaning they don’t break down and they can accumulate over time. In recent years, experts have become increasingly concerned by the potential effects of high concentrations of PFAS on human health.*

PFAS contamination in West Michigan has unfortunately made [headlines](https://www.wzzm13.com/article/news/local/judge-approves-54m-settlement-kent-co-residents-wolverine-3m-pfas-case/69-091ea42c-125a-4341-8ea0-238fdac4d06e), reminiscent of other environmental advocacy efforts like that of renowned whistleblower [Erin Brockovich](https://en.wikipedia.org/wiki/Erin_Brockovich).

There are multiple PFAS-related data sets available from the state of Michigan contained in a single zipped file on [GitHub](https://github.com/dilernia/STA418-518/blob/main/Data/pfas_data.zip) <https://github.com/dilernia/STA418-518/blob/main/Data/pfas_data.zip>:

1. **Surface water sampling data**; PFAS concentrations in surface water samples 🏞🐟

2. **Public water supply sampling data**; PFAS concentrations in municipal drinking water samples 💧🫗

3. **PFAS sites data** ; PFAS contamination or investigation sites of interest ☢️🏭

The original data dictionary for the full surface water data is available at <https://gis-egle.hub.arcgis.com/datasets/egle::pfas-surface-water-sampling/about>.

A non-comprehensive laboratory PFAS analyte list (showing for example that PFTeDA is Perfluorotetradecanoic acid) for analyzing data collected by Michigan’s Departments of Environment, Great Lakes, and Energy (EGLE), Health and Human Services (HHS), Agriculture and Rural Development, and Natural Resources can be found [here](https://www.michigan.gov/pfasresponse/-/media/Project/Websites/PFAS-Response/Sampling-Guidance/Minimum-Laboratory-Analyte-List.pdf?rev=a35aba56ec5a4922b986f01e25c1a19d&hash=04E6F164AA5F5CD29B83B39983341345): <https://www.michigan.gov/pfasresponse/-/media/Project/Websites/PFAS-Response/Sampling-Guidance/Minimum-Laboratory-Analyte-List.pdf?rev=a35aba56ec5a4922b986f01e25c1a19d&hash=04E6F164AA5F5CD29B83B39983341345>.

## Source

Data was obtained from the state of Michigan website <https://gis-egle.hub.arcgis.com/search?q=pfasgis>.

## Data Dictionary

### Surface water sampling data

```{r PFAS_surface_data_dictionary}
# Creating data dictionary
clean_surface_water_dictionary <- surface_water_dictionary |> 
  dplyr::rename(Variable = `Field Name`) |> 
  mutate(Variable = case_when(!str_starts(Variable, pattern = "CA") ~ make_clean_names(Variable),
                              TRUE ~ Variable)) |> 
  dplyr::filter(Variable %in% colnames(surface_water_long)) |> 
  bind_rows(tibble(Variable = c("geoid",
                                "analyte",
                                "analyte_value"),
                   Description = c("geographic region ID with the first 2 digits being the state Federal Information Processing Standard (FIPS) code \nand the last 3 digits the county FIPS code",
                                   "PFAS analyte",
                                   "The analyte concentration in the sample measured in parts per trillion (ppt or ng/L)"))) |> 
  mutate(Variable = factor(Variable, levels = colnames(surface_water_long))) |> 
  arrange(Variable)

# Printing data dictionary
clean_surface_water_dictionary |> 
  gt()
```

```{r PFAS_surface_explore}
#| include: false
#| eval: true
# Aggregating PFAS values for most recent measurement of each site
surface_water_total <- surface_water_long |> 
  dplyr::mutate(watershed = case_when(watershed == "Huron" ~ "Lake Huron",
                                      TRUE ~ watershed)) |> 
  group_by(geoid, longitude, latitude, site_code, location_code, watershed, waterbody) |> 
  slice_max(collection_date, n = 1) |> 
  group_by(geoid, longitude, latitude, collection_date, site_code, location_code, watershed, waterbody) |> 
  summarize(total_analytes = sum(analyte_value, na.rm = TRUE)) |> 
  ungroup()

# Ten highest PFAS locations
surface_water_total |> 
  slice_max(total_analytes, n = 10)

# Side-by-side box plots of total PFAS levels by Great Lake
surface_water_total |> 
  dplyr::filter(str_detect(watershed, pattern = "Huron|Lake"), 
                !str_detect(watershed, pattern = "Clair|Lone")) |> 
    ggplot(aes(y = fct_reorder(watershed, total_analytes, .fun = max),
               x = total_analytes)) +
    geom_boxplot() + 
    labs(title = "PFAS levels measured in the Great Lakes",
         x = "Total analyte level (ppt)",
         y = "") +
    ggthemes::theme_few()

# Distribution of PFAS levels measured in Kent County
surface_water_total |> 
  dplyr::mutate(county_fips = str_sub(geoid, start = 3, end = 5)) |> 
  dplyr::filter(county_fips == "081") |> 
ggplot(aes(x = total_analytes)) +
    geom_histogram(color = "black", fill = "dodgerblue") + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    labs(title = "PFAS levels measured in Kent County",
         x = "Total analyte level (ppt)",
         y = "Frequency") +
    ggthemes::theme_few()

# Occurrences of different analytes
analyte_summary <- surface_water_long |> 
    dplyr::mutate(watershed = case_when(watershed == "Huron" ~ "Lake Huron",
                                        TRUE ~ watershed)) |> 
    group_by(geoid, longitude, latitude, site_code, location_code, watershed, waterbody) |> 
    slice_max(collection_date, n = 1) |> 
    ungroup() |> 
    dplyr::filter(analyte_value > 0) |> 
    group_by(analyte) |> 
    summarize(min = min(analyte_value), 
              Q1 = quantile(analyte_value, probs = 0.25),
              median = quantile(analyte_value, probs = 0.5),
              Q2 = quantile(analyte_value, probs = 0.75),
              max = max(analyte_value), 
              total = sum(analyte_value),
              n_detected = n()) |> 
  arrange(desc(total))

# Ten highest analytes
analyte_summary |> 
  slice_max(total, n = 10)

# Number of unique analytes detected by Great Lake
surface_water_long |> 
    dplyr::mutate(watershed = case_when(watershed == "Huron" ~ "Lake Huron",
                                        TRUE ~ watershed)) |> 
  dplyr::filter(str_detect(watershed, pattern = "Huron|Lake"), 
                !str_detect(watershed, pattern = "Clair|Lone"),
                analyte_value > 0) |> 
  count(watershed, analyte) |> 
  count(watershed)
```

### Public water supply sampling data

From the [EGLE website](https://www.michigan.gov/egle/maps-data/mpart-pfas-gis):

*The public water supply results include information from both the State of Michigan funded drinking water Per- and Polyfluoroalkyl Substances (PFAS) sampling effort, and ongoing PFAS compliance monitoring of public water supplies in Michigan. It includes, but is not limited to, sample location, sample date, and sample results...*

*The combined public water supply results data is representative of the PFAS sampling locations, with each location having multiple samples taken, dates for each sample, and the analytical results for each sample.*

```{r PFAS_public_data_dictionary}
# Creating data dictionary
clean_public_water_dictionary <- surface_water_dictionary |> 
  dplyr::rename(Variable = `Field Name`) |> 
  mutate(Variable = case_when(!str_starts(Variable, pattern = "CA") ~ make_clean_names(Variable),
                              TRUE ~ Variable)) |> 
  dplyr::filter(Variable %in% colnames(public_water_long)) |> 
  bind_rows(tibble(Variable = c("geoid",
                                "system_name",
                                "system_type",
                                "sample_date",
                                "lab_name_code",
                                "analyte",
                                "analyte_value"),
                   Description = c("geographic region ID with the first 2 digits being the state Federal Information Processing Standard (FIPS) code \nand the last 3 digits the county FIPS code",
                                   "Name of water supply system",
                                   "Type of water supply system",
                                   "The date that the sample was collected",
                                   "Code for processing lab",
                                   "PFAS analyte",
                                   "The analyte concentration in the sample measured in parts per trillion (ppt or ng/L)"))) |> 
  mutate(Variable = factor(Variable, levels = colnames(public_water_long))) |> 
  arrange(Variable)

# Printing data dictionary
clean_public_water_dictionary |> 
  gt()
```

Handling data issues identified

```{r}
# Standardizing the variable names in the data sets
standardize_column_names <- function(df) {
  df |> janitor::clean_names()
}

# Apply to data sets
public_water_long <- public_water_long |> standardize_column_names()
surface_water_long <- surface_water_long |> standardize_column_names()
pfas_sites <- pfas_sites |> standardize_column_names()
```

```{r}
# Function to clean string columns (excluding numeric and analyte columns)
clean_strings_title_case <- function(df) {
  # Columns to clean
  cols_to_clean <- df %>%
    select(where(is.character)) %>%      
    select(-matches("analyte", ignore.case = TRUE)) %>%  
    colnames()
  
  # Applying title case to selected columns
  df %>%
    mutate(across(all_of(cols_to_clean), str_to_title))
}

# Application to datasets
public_water_long <- clean_strings_title_case(public_water_long)
surface_water_long <- clean_strings_title_case(surface_water_long)
pfas_sites <- clean_strings_title_case(pfas_sites)
```

```{r}
# Identify repeated measures (use later for trend analysis)
repeated_measures <- surface_water_long |>
  count(site_code, collection_date, visit_id, name = "n_per_visit") |>  # first count
  add_count(site_code, name = "n_per_site") |>                          # second count
  filter(n_per_site >= 2)

```

```{r}
public_water_long |> 
  distinct(system_name, latitude, longitude, sample_date)
```

```{r}
# Creating a function to create unique records in the datasets
get_max_concentration <- function(df, site_col, date_col, value_col) {
  
  df_total <- df |>
    group_by(across(all_of(c(site_col, date_col)))) |>
    summarise(total_concentration = sum(.data[[value_col]], na.rm = TRUE), .groups = "drop")
  
  df_max <- df_total |>
    group_by(across(all_of(site_col))) |>
    slice_max(order_by = total_concentration, n = 1, with_ties = FALSE) |>
    ungroup()
  
  df_max_full <- df_max |>
    left_join(df, by = setNames(c(site_col, date_col), c(site_col, date_col)))
  
  return(df_max_full)
}

```

```{r}
# Application to surface_water_long
surface_max_full <- get_max_concentration(
  df = surface_water_long,
  site_col = "site_code",
  date_col = "collection_date",
  value_col = "analyte_value"
)

# Application to public_water_long
public_max_full <- get_max_concentration(
  df = public_water_long,
  site_col = "system_name",
  date_col = "sample_date",
  value_col = "analyte_value"
)
```


### PFAS sites

This data was last updated on 02/05/2025, and was downloaded on 02/25/2025 from the [EGLE](https://gis-egle.hub.arcgis.com/maps/17b26cf283624bf49705741e81fde0c4) website. Cleaned data in a more convenient CSV format is available [here](https://raw.githubusercontent.com/dilernia/STA418-518/refs/heads/main/Data/pfas_sites.csv): <https://raw.githubusercontent.com/dilernia/STA418-518/refs/heads/main/Data/pfas_sites.csv>. From the EGLE website:

*The Michigan PFAS Sites layer is the official list of PFAS sites in Michigan. This data is updated as new PFAS sites are encountered.*

The original website for the definitions of the site variables is [here](https://gis-egle.hub.arcgis.com/datasets/egle::michigan-pfas-sites/about): <https://gis-egle.hub.arcgis.com/datasets/egle::michigan-pfas-sites/about>.

```{r PFAS_sites_data_dictionary}
pfas_sites_descriptions <- c(
  "Geographic region ID with the first 2 digits being the state Federal Information Processing Standard (FIPS) code \nand the last 3 digits the county FIPS code",
  "Longitude coordinate of sample location",
  "Latitude coordinate of sample location",
  "Name of the site",
  "County where the site is located",
  "Physical address of the project site",
  "City where the project is physically located",
  "Zip code where the project is physically located",
  "Describes the kind of site by type",
  "Indicates whether residential wells have, or will be sampled (Domains = Yes, No, TBD, N/A)",
  "Name of contact person",
  "Contact person email",
  "Contact person phone number",
  "Web address to the site investigation summary",
  "Geographic location description of the site",
  "Indicates if the site is associated with the military",
  "Date the site was published on the Michigan PFAS Action Response Team website",
  "Background information about the site",
  "Information about drinking water quality and safety",
  "The state of Michigan's planned or upcoming activities related to the site"
)

# Printing data dictionary
pfas_sites |> 
  data_dictionary(descripts = pfas_sites_descriptions)
```

```{r}
library(skimr)
# Exploring the data set using skim package
# Surface water dataset
skim(surface_max_full)

```

```{r}
# Public water supply dataset
skim(public_max_full)
```

```{r}
# PFAS sites dataset
skim(pfas_sites)
```

Exploring Missingness in data sets and Exploartory Data Analysis

```{r, warning=FALSE}
library(tidyverse)   
library(sf)
library(patchwork)
library(janitor)
library(naniar)
library(skimr)
library(ggrepel)
library(ggthemes)
library(scales)
library(reshape2)
library(ggcorrplot)
library(flextable)
```

```{r}
# % Missingness in surafce water data
gg_miss_var(surface_max_full, show_pct = TRUE) +
  labs(title = "Missingness by Variable: PFAS Surface Water Data",
       caption = "Data source: Michigan PFAS Action Response Team (MPART), EGLE",
       x = "Variable Name",
       y = "% Missingness")

# % Missingness in public water
gg_miss_var(public_max_full, show_pct = TRUE) +
  labs(title = "Missingness by Variable: PFAS Surface Water Data",
       caption = "Data source: Michigan PFAS Action Response Team (MPART), EGLE",
       x = "Variable Name",
       y = "% Missingness")

# % Missingness in PFAS sites
gg_miss_var(pfas_sites, show_pct = TRUE) +
  labs(title = "Missingness by Variable: PFAS Surface Water Data",
       caption = "Data source: Michigan PFAS Action Response Team (MPART), EGLE",
       x = "Variable Name",
       y = "% Missingness")

```

Exploratory Data Analysis
*Analytes analysis in both data sets*

```{r}
# Unique analytes in public water dataset
unique(public_max_full$analyte)
unique_public
```

```{r}
# Unique analytes in surface water dataset
unique(surface_max_full$analyte)
unique_surface
```

*Surface Water Visualizations*

```{r}
# Bar plots showing the surface water PFAS concentration by top and bottom 15 counties
# Load county shapefile
counties_sf <- st_read("cb_2018_us_county_500k.shp") |>
  st_drop_geometry() |>
  select(GEOID, NAME)

# Aggregate total PFAS by county
pfas_county_summary <- surface_water_long |>
  group_by(geoid) |>
  summarize(total_value = sum(analyte_value, na.rm = TRUE)) |>
  ungroup() |>
  left_join(counties_sf, by = c("geoid" = "GEOID")) |>
  rename(NAME = NAME)

# Top 15 counties
top15 <- pfas_county_summary |>
  slice_max(total_value, n = 15) |>
  mutate(group = "Top 15")

# Bottom 15 counties
bottom15 <- pfas_county_summary |>
  slice_min(total_value, n = 15) |>
  mutate(group = "Bottom 15")

# Combine top and bottom counties
pfas_top_bottom <- bind_rows(top15, bottom15)

# Plot
pfas_top_bottom |>
  ggplot(aes(x = fct_reorder(NAME, total_value), 
             y = total_value, 
             fill = group)) +
  geom_col() +
  geom_text(aes(label = round(total_value, 1)), 
            hjust = -0.05, size = 3.2) +
  coord_flip() +
  scale_fill_manual(values = c("Top 15" = "mediumseagreen", "Bottom 15" = "orange")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  facet_wrap(~group, scales = "free_y") +
  labs(
    title = "Surface Water PFAS by Top and Bottom 15 Counties",
    x = "County",
    y = "Mean PFAS (ng/L)",
    caption = "Data source: Michigan PFAS Action Response Team (MPART), EGLE"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.text.x = element_blank(),  # removes numeric tick labels
    axis.text.y = element_text(face = "bold"), # keeps county names visible
    plot.caption = element_text(face = "bold"),
    legend.position = "none"
  )
```

```{r}
# Top 10 analyte category by total % & absolute concentration

surface_water_long |>
  filter(analyte_value > 0) |>
  group_by(analyte) |>
  summarize(total_concentration = sum(analyte_value, na.rm = TRUE)) |>
  ungroup() |>
  mutate(share = total_concentration / sum(total_concentration)) |>
  slice_max(share, n = 10) |>
  ggplot(aes(x = fct_reorder(analyte, share), 
             y = share,
             fill = share)) +
  geom_col() +
  geom_text(aes(
    label = paste0(
      percent(share, accuracy = 0.1), 
      " (", comma(round(total_concentration, 1)), " ng/L)"
    )),
    hjust = -0.05, size = 3.2
  ) +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.15))) +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 20)) +
  scale_fill_viridis_c(option = "cividis", direction = -1) +
  labs(title = "Top 10 PFAS Analytes in Surface Water",
       x = "Analyte_Category", 
       y = "PFAS Concentration per Analyte",
       fill = "Share",
       caption = "Data source: Michigan PFAS Action Response Team (MPART), EGLE") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 15),
        axis.title.y = element_text(face = "bold"),
        axis.title.x = element_text(face = "bold"),
        axis.text.x = element_blank(),   # removes the % tick labels
        plot.caption = element_text(face = "bold"),
        legend.position = "none")
```

```{r}
# Map of total surface water PFAS by county
# Load county shapefile
mi_counties <- st_read("cb_2018_us_county_500k.shp") |>
  filter(STATEFP == "26")   

# Aggregate public water PFAS by county
public_county_map <- public_water_long |>
  group_by(geoid) |>
  summarize(total_value = sum(analyte_value, na.rm = TRUE), .groups = "drop") |>
  left_join(mi_counties, by = c("geoid" = "GEOID")) |>
  st_as_sf()

ggplot(public_county_map) +
  geom_sf(aes(fill = total_value), color = "white") +
  scale_fill_viridis_c(option = "magma", na.value = "grey90") +
  labs(
    title = "Total PFAS Concentration in Public Water by County",
    fill = "Total PFAS (ng/L)",
    caption = "Data source: Michigan PFAS Action Response Team (MPART), EGLE"
  ) +
  theme_minimal()

```
*Public Water Analysis*
```{r}
# Michigan PFAS drinking water standards (ppt / ng/L)
str(public_max_full$total_concentration)
```

```{r}
names(public_max_full)
```


```{r}
# Top 10 public water systems by total PFAS
public_water_top10 <- public_water_long |>
  group_by(system_name) |>
  summarize(total_value = sum(analyte_value, na.rm = TRUE)) |>
  ungroup() |>
  slice_max(order_by = total_value, n = 10)

# Improved bar plot
ggplot(public_water_top10, aes(x = fct_reorder(system_name, total_value), 
                               y = total_value, 
                               fill = total_value)) +
  geom_col() +
  geom_text(aes(label = round(total_value, 2)), hjust = -0.1, size = 3.5) +  
  coord_flip() +
  scale_fill_viridis_c(option = "viridis", direction = -1) +  
  scale_x_discrete(labels = function(x) str_wrap(x, width = 25)) +  
  labs(
    title = "Top 10 Public Water Systems by Total PFAS",
    x = "Public Water System",
    y = "Total PFAS (ng/L)",
    fill = "Total PFAS",
    caption = "Data source: Michigan PFAS Action Response Team (MPART), EGLE"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    legend.position = "none"
  ) +
  expand_limits(y = max(public_water_top10$total_value) * 1.1)
 

```

```{r}
# Top 10 PFAS analytes in public water
top_analytes <- public_water_long |>
  filter(analyte_value > 0) |>
  group_by(analyte) |>
  summarize(total_concentration = sum(analyte_value, na.rm = TRUE), .groups = "drop") |>
  slice_max(total_concentration, n = 10)

ggplot(top_analytes, aes(x = fct_reorder(analyte, total_concentration), 
                         y = total_concentration, 
                         fill = total_concentration)) +
  geom_col() +
  geom_text(aes(label = comma(round(total_concentration, 1))), 
            hjust = -0.05, size = 3.5) +  # labels outside bars
  coord_flip() +
  scale_fill_viridis_c(option = "cividis", direction = -1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +  # wrap long names
  labs(
    title = "Top 10 PFAS Analytes in Public Water Systems",
    x = "PFAS Analyte",
    y = "Total PFAS Concentration (ng/L)",
    fill = "Total Concentration",
    caption = "Data source: Michigan PFAS Action Response Team (MPART), EGLE"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    legend.position = "none"
  ) +
  expand_limits(y = max(top_analytes$total_concentration) * 1.1)  # space for labels
```

```{r}
# Map of total public water PFAS by county
public_county_map <- public_water_long |>
  group_by(geoid) |>
  summarize(total_value = sum(analyte_value, na.rm = TRUE), .groups = "drop") |>
  left_join(mi_counties, by = c("geoid" = "GEOID")) |>
  st_as_sf()

ggplot(public_county_map) +
  geom_sf(aes(fill = total_value), color = "white") +
  scale_fill_viridis_c(option = "magma", na.value = "grey90") +
  labs(
    title = "Total PFAS Concentration in Public Water by County",
    x = "Longitude",
    y = "Latitude",
    fill = "Total PFAS (ng/L)",
    caption = "Data source: Michigan PFAS Action Response Team (MPART), EGLE"
  ) +
  theme_minimal()


```

*PFAS sites Analysis*

```{r}
# Top 10 counties by PFAS site count
sites_top10_counties <- pfas_sites |>
  count(county, sort = TRUE) |>
  slice_max(n, n = 10)

ggplot(sites_top10_counties, aes(x = fct_reorder(county, n), y = n, fill = n)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.02, size = 3.5) +  
  coord_flip() +
  scale_fill_viridis_c(option = "viridis", direction = -1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +  
  labs(
    title = "Counties with Highest PFAS sites",
    x = "County",
    y = "Number of PFAS Sites",
    fill = "Site Count",
    caption = "Data source: Michigan PFAS Action Response Team (MPART), EGLE"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    legend.position = "none"
  )

```


```{r}
# Map of PFAS sites across Michigan
pfas_sites_sf <- pfas_sites |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  mutate(type = str_replace_all(type, "mlitary", "military"),
         type = str_to_title(type)) 
ggplot() +
  geom_sf(data = mi_counties, fill = "grey90", color = "white") +
  geom_sf(data = pfas_sites_sf, aes(color = type), size = 1.5, alpha = 0.7) +
  scale_color_viridis_d() +
  labs(
    title = "PFAS Sites Across Michigan",
    color = "Site Type",
    caption = "Data source: Michigan PFAS Action Response Team (MPART), EGLE"
  ) +
  theme_minimal()
```

```{r}
# Number of PFAS sites per type
pfas_sites |>
  mutate(type = str_replace_all(type, "mlitary", "military"),
         type = str_to_title(type)) |>
  count(type, sort = TRUE) |>
  ggplot(aes(x = fct_reorder(type, n), y = n, fill = n)) +
  geom_col() +
  geom_text(aes(label = n), 
            hjust = -0.05,  # slightly outside the bar
            size = 3.5) +
  coord_flip() +
  scale_fill_viridis_c(option = "cividis") +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +  # add 15% space at top
  labs(
    title = "Number of PFAS Sites by Type",
    x = "Site Type",
    y = "Number of Sites",
    fill = "Count",
    caption = "Data source: Michigan PFAS Action Response Team (MPART), EGLE"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

*Combined Exploratory Data Analysis*

```{r, warning=FALSE}
library(tidytext)

# Read county shapefile
counties_sf <- st_read("cb_2018_us_county_500k.shp") |>
  st_drop_geometry() |>
  select(GEOID, NAME)

# Convert GEOID to character for joining
pfas_county_summary <- pfas_county_summary |> mutate(geoid = as.character(geoid))
public_county_totals <- public_water_long |>
  group_by(geoid) |>
  summarize(total_public = sum(analyte_value, na.rm = TRUE), .groups = "drop") |>
  mutate(geoid = as.character(geoid))
sites_county <- pfas_sites |>
  count(geoid) |>
  rename(n_sites = n) |>
  mutate(geoid = as.character(geoid))

# Combine totals by county
county_combined <- pfas_county_summary |>
  select(geoid, total_value) |>
  rename(total_surface = total_value) |>
  full_join(public_county_totals, by = "geoid") |>
  full_join(sites_county, by = "geoid")

# Summary table with county names
county_summary <- county_combined |>
  left_join(counties_sf, by = c("geoid" = "GEOID")) |>
  relocate(NAME, .before = geoid) |>
  select(NAME, total_surface, total_public, n_sites)

# Flextable for top 10 counties by total_surface
top10_county_table <- county_summary |> slice_max(total_surface, n = 10)
flextable(top10_county_table) |> autofit()

# Pivot longer for faceted bar plot
county_long <- top10_county_table |>
  pivot_longer(cols = c(total_surface, total_public, n_sites),
               names_to = "metric", values_to = "value") |>
  mutate(NAME = reorder_within(NAME, value, metric))  # reorder within facet

# Custom labels for facets
metric_labels <- c(
  total_surface = "Total Surface PFAS (ng/L)",
  total_public = "Total Public PFAS (ng/L)",
  n_sites = "Number of PFAS Sites"
)

# Faceted bar plot
county_long |>
  ggplot(aes(x = NAME, y = value, fill = metric)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = round(value, 2)), 
            hjust = -0.1, size = 4, color = "black") +
  facet_wrap(~ metric, scales = "free_y", labeller = labeller(metric = metric_labels)) +
  coord_flip(clip = "off") +  
  scale_x_reordered() +  # handles reordered factor levels
  scale_fill_manual(values = c("steelblue", "tomato", "darkgreen")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)), breaks = NULL) + 
  labs(x = "County",
       y = NULL,
       title = "Top 10 Counties by Metric: Surface & Public PFAS and Number of Sites",
       caption = "Data source: Michigan PFAS Action Response Team (MPART), EGLE") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.caption = element_text(size = 10, face = "bold"),
    axis.text.x = element_blank(),
    axis.text.y = element_text(face = "bold")
  )


```

```{r}
# Total PFAS concentration by water type
total_pfas <- bind_rows(
  surface_water_long |> summarize(total = sum(analyte_value, na.rm = TRUE)) |> mutate(water_type = "Surface Water"),
  public_water_long  |> summarize(total = sum(analyte_value, na.rm = TRUE)) |> mutate(water_type = "Public Water")
) |> 
  mutate(pct = total / sum(total) * 100,
         lbl = paste0(water_type, " (", round(pct, 1), "%)"))

# Pie chart
total_pfas |>
  ggplot(aes(x = "", y = total, fill = water_type)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = lbl), position = position_stack(vjust = 0.5), size = 5) +
  scale_fill_manual(values = c("Surface Water" = "steelblue", "Public Water" = "tomato")) +
  labs(title = "Total PFAS Concentration by Water Type",
       caption = "Data source: Michigan PFAS Action Response Team (MPART), EGLE") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 16),
        plot.caption = element_text(face = "bold"))

```
