---
title: "Public Water Hazard Index Analysis"
format: docx
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
set.seed(1994)

# Loading necessary packages
library(tidyverse)
library(naniar)
library(janitor)
library(sf)
library(gt)
library(skimr)
library(tidycensus)
library(maps)
library(patchwork)
library(ggrepel)
library(ggthemes)
library(scales)
library(reshape2)
library(ggcorrplot)
library(flextable)
library(dplyr)
library(tidyr)
library(plotly)
library(leaflet)
library(reactable)
library(reactablefmtr)
library(htmltools)
library(leaflet.extras2)
library(leaflegend)
library(htmlwidgets)
library(leaflet.extras)
library(crosstalk)
library(stringr)
library(forcats)
library(webshot2)
library(svglite)



# Function for creating data dictionary
data_dictionary <- function(myData, descripts) {
tibble(Variable = colnames(myData),
       Type = map_chr(myData, .f = function(x){typeof(x)[1]}),
       Description = descripts) |> 
  gt()
}
```

```{r PFAS_packages}
#| eval: true
#| include: false

# Create function to add geoid variable given longitude and latitude for county-level data
add_geoid <- function(my_data) {
  # Create sf object to help add geoid column
my_data_sf <- my_data |> 
st_as_sf(coords = c("longitude", "latitude"))

# Set coordinate reference system (CRS) to match counties data
counties <- st_read("GeoFiles//cb_2018_us_county_500k.shp")
st_crs(my_data_sf) <- st_crs(counties)

ret_data <- my_data_sf |> 
  st_join(counties) |> 
    mutate(longitude = st_coordinates(geometry)[, 1],
           latitude = st_coordinates(geometry)[, 2]) |> 
  st_drop_geometry() |> 
  as_tibble() |> 
  dplyr::select(GEOID, longitude, latitude) |> 
  dplyr::rename(geoid = GEOID) |> 
  distinct() |> 
  right_join(my_data)

return(ret_data)
}
```

```{r PFAS_surface_clean}
#| eval: true
#| include: false

# Importing data on PFAS levels in surface water in Michigan
analyte_info <- read_csv("PFAS_Surface_Water_Sampling.csv") |> 
              dplyr::select(-c(X:Latitude, GlobalID, OBJECTID,
                               ends_with("Flag")))

surface_water <- read_csv("PFAS_Surface_Water_Sampling.csv") |> 
  clean_names() |> 
  dplyr::select(-ends_with("_flag")) |> 
  dplyr::select(-c(cas307244_pf_hx_a:cas919005144_adona_rl)) |> 
  bind_cols(analyte_info)

# Importing data on surface water data dictionary
surface_water_dictionary <- read_csv("surface_water_data_dictionary.csv")

# Creating long version of surface water data
surface_water_long <- surface_water |> 
  dplyr::rename(object_id = objectid) |> 
  dplyr::select(longitude, latitude, object_id, everything(), 
                -c(x:y), -ends_with("Flag"), -global_id) |> 
  pivot_longer(cols = colnames(analyte_info), 
               names_to = "analyte", 
               values_to = "analyte_value") |> 
  dplyr::select(-matrix, -unit, -object_id) |> 
  add_geoid()

# Saving to external CSV
write_csv(surface_water_long, file = "pfas_surface_water_long.csv")
```

```{r PFAS_public_clean}
#| eval: true
#| include: false

# Importing shape file data for public water supply PFAS levels data
public_water_shape <- st_read("Public_Water_Supply_Sampling_Hexbins_and_Results/Public_Water_Supply_Sampling_Hexbins.shp") |> 
  mutate(centroid = st_centroid(geometry),
         longitude = st_coordinates(centroid)[, 1],
         latitude = st_coordinates(centroid)[, 2]) |> 
  dplyr::select(HexID, longitude, latitude) |> 
  st_drop_geometry() |> 
  as_tibble()

# Importing data on public water supply PFAS levels
public_water <- read_csv("Public_Water_Supply_Sampling_Hexbins_and_Results.csv")

# Adding geoid, longitude, and latitude information to public water data
# and cleaning variable names
public_water_wide <- public_water |> 
  left_join(public_water_shape, by = "HexID") |> 
  dplyr::select(-ends_with(c("Result", "Flags"))) |> 
  clean_names() |> 
  dplyr::rename(object_id = objectid) |> 
  dplyr::select(object_id:sys_loc_code, longitude, latitude, everything()) |> 
  bind_cols(dplyr::select(public_water, ends_with("Result"))) |> 
  add_geoid()

# Pivoting public water data from wide to long format
public_water_long <- public_water_wide |> 
  pivot_longer(cols = c(ends_with(c("Result"))), 
               names_to = "analyte", 
               values_to = "analyte_value") |> 
  mutate(analyte = gsub(analyte, pattern = "Result", replacement = "")) |> 
  dplyr::select(-c(hex_id, 
                   wssn, loc_name:sys_loc_code, 
                   phase_code:task_type,
                   treatment_status:sys_sample_code,
                   lab_number:lab_sdg, 
                   object_id, position_source,
                   analytical_method,
                   sampling_results_count)) |> 
  mutate(system_type = fct_recode(system_type,
                               "Non-Community Water Supply (Adult Foster Care Provider)" = "ADFSTC",
                               "Non-Community Water Supply (Children's Camp)" = "CHLCMP",
                               "Non-Community Water Supply (Child Care Provider)" = "DAYCARE",
                               "Non-Community Water Supply (Industry)" = "INDUS",
                               "Non-Community Water Supply (Medical Care Provider)" = "MEDCAR",
                               "Non-Community Water Supply (Hotel or Motel)" = "MOTEL",
                               "Community Water Supply (for example Municipal Supply, Apartment, Nursing Home, Prison, etc.)" = "MUN",
                               "Office Building" = "OFFICE",
                               "Park" = "PARK",
                               "Residential" = "RESD",
                               "School" = "SCH",
                               "Tribal Lands" = "TRB"))


# Saving to external CSV
write_csv(public_water_long, file = "pfas_public_water_long.csv")
```

```{r PFAS_sites_import}
#| eval: true
#| include: false

pfas_site_path <- "pfas_sites.csv"

if(file.exists(pfas_site_path) == FALSE) {
# Importing data on identified sites from https://www.michigan.gov/pfasresponse
# Downloaded on 02/25/2025
pfas_sites <- read_csv("Michigan_PFAS_Sites.csv") |> 
  janitor::clean_names() |> 
  dplyr::select(facility:site_type, location:military, facility_date, site_background:anticipated_activities, -tier, -site_type) |> 
  add_geoid()

# Saving cleaned data
write_csv(pfas_sites, file = pfas_site_path)
} else {
  # Importing
  pfas_sites <- read_csv(pfas_site_path)
}
```

```{r}
# Drop flag columns, replace NA in PFAS Result columns with 0, and ensure numeric
public_water_wide <- public_water_wide |> 
  select(-ends_with("Flags")) |> 
  mutate(across(ends_with("Result"), ~replace_na(as.numeric(.x), 0)))

# Calculating the hazard index
public_water_wide <- public_water_wide |>
  mutate(Hazard_Index = (HFPODAResult / 10) +
         (PFNAResult / 10) +
         (PFBSResult / 2000) +
         (PFHxSResult / 10))
```

```{r}
# Displaying the high index areas with the highest analyte contributor in a table
high_hazard <- public_water_wide |> 
  filter(Hazard_Index > 1) |> 
  select(hex_id, sample_date, Hazard_Index, HFPODAResult, PFNAResult, PFBSResult, PFHxSResult) |> 
  arrange(desc(Hazard_Index)) |> 
  flextable() |> 
  set_caption("Public Water Locations with Hazard Index > 1") |> 
  autofit()
```

```{r}
# Extracting the maximum hazard per hex ID
public_water_max <- public_water_wide |> 
  group_by(hex_id) |> 
  slice_max(order_by = Hazard_Index, n = 1, with_ties = FALSE) |> 
  ungroup()
```


```{r}
# Changing the wide format to long format
public_water_max_long <- public_water_max |>
pivot_longer(cols = ends_with("Result"),   
               names_to = "analyte",
               values_to = "analyte_value") |>
mutate(analyte = gsub("Result$", "", analyte))
```


```{r}
merged_data <- public_water_max_long |>
  left_join(
    pfas_sites |> mutate(geoid = as.character(geoid)), 
    by = "geoid",
    relationship = "many-to-many"
  )
```

```{r}
# Reusable JS for auto-hover
js_auto_hover <- "
function(el, x) {
  // Determine number of points depending on chart orientation
  var d = el.data[0];
  var totalPoints = 0;

  if (d.x && d.x.length > 0) {
    totalPoints = d.x.length;   // vertical bars or line charts
  } 
  if (d.y && d.y.length > 0) {
    totalPoints = d.y.length;   // horizontal bars
  }

  var i = 0;
  setInterval(function() {
    Plotly.Fx.hover(el, [{ curveNumber: 0, pointNumber: i }], 'xy');
    i = (i + 1) % totalPoints;
  }, 2000);
}
"
```

```{r}
# Reusable JS for auto-hover on pie charts
js_auto_hover_pie <- "
function(el, x) {
  var totalSlices = el.data[0].labels.length; // number of pie slices
  var i = 0;
  setInterval(function() {
    Plotly.Fx.hover(el, [{ curveNumber: 0, pointNumber: i }]);
    i = (i + 1) % totalSlices;
  }, 2000);
}
"
```

```{r}
# Prepare system type mapping
system_type_map <- tibble(
  short_name = c(
    "ADFSTC", "CHLCMP", "DAYCARE", "INDUS", "MEDCAR",
    "MOTEL", "MUN", "OFFICE", "PARK", "RESD", "SCH", "TRB"
  ),
  full_name = c(
    "Non-Community Water Supply (Adult Foster Care Provider)",
    "Non-Community Water Supply (Children's Camp)",
    "Non-Community Water Supply (Child Care Provider)",
    "Non-Community Water Supply (Industry)",
    "Non-Community Water Supply (Medical Care Provider)",
    "Non-Community Water Supply (Hotel or Motel)",
    "Community Water Supply (for example Municipal Supply, Apartment, Nursing Home, Prison, etc.)",
    "Office Building",
    "Park",
    "Residential",
    "School",
    "Tribal Lands"
  )
)

# Apply short names to the dataset
public_water_max_long <- public_water_max_long |>
  mutate(system_short = recode(system_type,
    "Non-Community Water Supply (Adult Foster Care Provider)" = "ADFSTC",
    "Non-Community Water Supply (Children's Camp)" = "CHLCMP",
    "Non-Community Water Supply (Child Care Provider)" = "DAYCARE",
    "Non-Community Water Supply (Industry)" = "INDUS",
    "Non-Community Water Supply (Medical Care Provider)" = "MEDCAR",
    "Non-Community Water Supply (Hotel or Motel)" = "MOTEL",
    "Community Water Supply (for example Municipal Supply, Apartment, Nursing Home, Prison, etc.)" = "MUN",
    "Office Building" = "OFFICE",
    "Park" = "PARK",
    "Residential" = "RESD",
    "School" = "SCH",
    "Tribal Lands" = "TRB"
  ))
```

## PFAS Concentration

```{r PFAS_surface_explore}
#| include: false
#| eval: true
# Aggregating PFAS values for most recent measurement of each site
surface_water_total <- surface_water_long |> 
  dplyr::mutate(watershed = case_when(watershed == "Huron" ~ "Lake Huron",
                                      TRUE ~ watershed)) |> 
  group_by(geoid, longitude, latitude, site_code, location_code, watershed, waterbody) |> 
  slice_max(collection_date, n = 1) |> 
  group_by(geoid, longitude, latitude, collection_date, site_code, location_code, watershed, waterbody) |> 
  summarize(total_analytes = sum(analyte_value, na.rm = TRUE)) |> 
  ungroup()

```

```{r}
# Standardizing the variable names in the data sets
standardize_column_names <- function(df) {
  df |> janitor::clean_names()
}

# Apply to data sets
public_water_long_conc <- public_water_long |> standardize_column_names()
surface_water_long <- surface_water_long |> standardize_column_names()
pfas_sites_conc <- pfas_sites |> standardize_column_names()
```

```{r}
# Function to clean string columns (excluding numeric and analyte columns)
clean_strings_title_case <- function(df) {
  # Columns to clean
  cols_to_clean <- df %>%
    select(where(is.character)) %>%      
    select(-matches("analyte", ignore.case = TRUE)) %>%  
    colnames()
  
  # Applying title case to selected columns
  df %>%
    mutate(across(all_of(cols_to_clean), str_to_title))
}

# Application to datasets
public_water_long_conc <- clean_strings_title_case(public_water_long)
surface_water_long <- clean_strings_title_case(surface_water_long)
pfas_sites_conc <- clean_strings_title_case(pfas_sites)
```

```{r}
# Function to get maximum analyte value per site
get_max_concentration <- function(df, site_col, analyte_col, value_col) {
  df_max <- df |>
    # Step 0: Replace NAs in analyte and value columns with 0
    mutate(
      !!analyte_col := ifelse(is.na(.data[[analyte_col]]), 0, .data[[analyte_col]]),
      !!value_col := ifelse(is.na(.data[[value_col]]), 0, .data[[value_col]])
    ) |>
    
    # Step 1: Group by site and analyte
    group_by(across(all_of(c(site_col, analyte_col)))) |>
    
    # Step 2: Select the row with the highest analyte value within each group
    slice_max(order_by = .data[[value_col]], n = 1, with_ties = FALSE) |>
    
    # Step 3: Ungroup
    ungroup()

  return(df_max)
}

```

```{r}
# Application in surface water
surface_max_full <- get_max_concentration(
  df = surface_water_long,
  site_col = "site_code",
  analyte_col = "analyte",
  value_col = "analyte_value"
)

# Application in public water
public_max_full <- get_max_concentration(
  df = public_water_long_conc,
  site_col = "system_name",
  analyte_col = "analyte",
  value_col = "analyte_value"
)
```

Poster Workflow

## Hazard Index

The hazard index is a unitless metric from the Environmental Protection Agency (EPA) that summarizes the adverse health effects associated with exposure to mixtures of the PFAS analytes PFNA, PFHxS, PFBS, and HFPO-DA. 

$$\text{Hazard Index} = \left(\frac{\text{HFPODA ppt}}{10 \text{ ppt}}\right) + \left(\frac{\text{PFBS ppt}}{2000 \text{ ppt}}\right) + \left(\frac{\text{PFNA ppt}}{10 \text{ ppt}}\right) + \left(\frac{\text{PFHxS ppt}}{10 \text{ ppt}}\right) $$

The EPA’s Hazard Index Maximum Contaminant Level (MCL) is set at 1, meaning there are potential elevated health risks for drinking water with hazard index levels exceeding 1.


```{r}
# Define main hazard analytes and colors
hazard_analytes <- c("HFPODA", "PFNA", "PFBS", "PFHxS")
hazard_colors <- c(
  "HFPODA" = "#332288",
  "PFNA"   = "#117733",
  "PFBS"   = "#44AA99",
  "PFHxS"  = "#88CCEE"
)

# Global theme for all plots
theme_set(
  theme_classic(base_size = 22) +
    theme(
      plot.title = element_text(face = "bold", color = "black"),
      axis.title = element_text(face = "bold", color = "black"),
      axis.text = element_text(face = "bold", color = "black"),
      legend.title = element_text(face = "bold", size = 22, color = "black"),
      legend.text  = element_text(face = "bold", size = 20, color = "black"),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    )
)
```

```{r}
#| label: analyte contribution
# Aggregate analyte-level hazard contributions
total_analyte_hazard <- public_water_max_long |>
  filter(analyte %in% hazard_analytes) |>           
  mutate(analyte_hazard = case_when(
    analyte == "HFPODA" ~ analyte_value / 10,
    analyte == "PFNA"   ~ analyte_value / 10,
    analyte == "PFBS"   ~ analyte_value / 2000,
    analyte == "PFHxS"  ~ analyte_value / 10
  )) |>
  group_by(analyte) |>
  summarise(
    total_contribution = sum(analyte_hazard, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_contribution)) |>
  mutate(
    percent = round(total_contribution / sum(total_contribution) * 100, 1),
    analyte = factor(analyte, levels = analyte)
  )

# Bar chart for analytes
p_bar <- ggplot(total_analyte_hazard, 
                aes(x = analyte, y = total_contribution, fill = analyte)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(percent, "%")), 
            vjust = -0.5, size = 7, fontface = "bold") +
  scale_fill_manual(values = hazard_colors) +
  labs(
    title = "Decomposed Hazard Index Contributions by Analyte",
    x = "Analyte",
    y = "Total Hazard Index Contribution"
  ) +
  ylim(0, max(total_analyte_hazard$total_contribution) * 1.15)

ggsave("analyte_contributions.svg", p_bar, width = 12, height = 8, dpi = 300)

```

```{r}
# Load Michigan counties
mi_counties <- st_read("GeoFiles/cb_2018_us_county_500k.shp", quiet = TRUE) %>%
  filter(STATEFP == "26") %>%
  st_transform(4326)

# Bounding box for both maps
mi_bbox <- st_bbox(mi_counties)
```

```{r}
#| label: hazard map
# Prepare analyte-wide data for hazard sites
analyte_wide <- public_water_max_long %>%
  filter(analyte %in% hazard_analytes) %>%
  select(hex_id, analyte, analyte_value) %>%
  pivot_wider(
    names_from = analyte,
    values_from = analyte_value,
    values_fill = 0
  )

# Prepare hazard sites
hazard_sites_full <- public_water_max_long %>%
  filter(Hazard_Index > 1) %>%
  distinct(hex_id, system_name, system_type, Hazard_Index, longitude, latitude) %>%
  rename(type = system_type) %>%
  left_join(analyte_wide, by = "hex_id") %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  mutate(
    Hazard_Index = as.numeric(Hazard_Index),
    system_name = tolower(system_name),
    type = tolower(type)
  )

# Hazard_map
map_plot2 <- ggplot() +
    geom_sf(data = mi_counties,
            fill = "white", color = "black", linewidth = 0.3) +
    geom_sf(data = hazard_sites_full,
            aes(size = Hazard_Index, fill = Hazard_Index),
            shape = 21, color = "black", stroke = 0.4,
            alpha = 0.8) +
    scale_fill_distiller(name = "Hazard Index", palette = "YlGnBu",
                         direction = 1, trans = "log1p",
                         breaks = breaks_log(n = 6)) +
    scale_size_continuous(name = "Hazard Index", trans = "log1p",
                          breaks = breaks_log(n = 6),
                          range = c(4, 18)) +
    coord_sf(
      xlim = c(mi_bbox["xmin"], mi_bbox["xmax"]),
      ylim = c(mi_bbox["ymin"], mi_bbox["ymax"]),
      expand = FALSE
    ) +
    guides(fill = guide_legend(),
           size = guide_legend()) +
    theme_map(base_size = 32) +
    theme(
        legend.position = c(0.02, 0.1),
        legend.justification = c("left", "bottom"),
        legend.key.size = unit(1.2, "cm"),
        legend.spacing.x = unit(0.4, "cm"),
        plot.margin = ggplot2::margin(0, 0, 0, 0),
        text = element_text(family = "Arial"),
        legend.title = element_text(face = "bold", size = 22, color = "black"),
        legend.text  = element_text(face = "bold", size = 20, color = "black")
    )

ggsave("hazard_map2.svg", map_plot2, width = 15, height = 18)
```

```{r}
#| label: hazard analyte by system_name
# System-level hazard decomposition
system_analytes <- public_water_max_long |>
  filter(
    Hazard_Index > 1,
    !is.na(analyte),
    analyte %in% hazard_analytes,
    !is.na(analyte_value),
    !is.na(system_name)
  ) |>
  mutate(
    system_name = tolower(system_name),
    analyte_hazard = case_when(
      analyte == "HFPODA" ~ analyte_value / 10,
      analyte == "PFNA"   ~ analyte_value / 10,
      analyte == "PFBS"   ~ analyte_value / 2000,
      analyte == "PFHxS"  ~ analyte_value / 10
    )
  ) |>
  group_by(system_name, analyte) |>
  summarise(
    total_hazard_analyte = sum(analyte_hazard, na.rm = TRUE),
    .groups = "drop"
  )

system_total <- system_analytes |>
  group_by(system_name) |>
  summarise(
    total_hazard = sum(total_hazard_analyte, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_hazard)) |>
  slice_head(n = 10)

system_analytes_top <- system_analytes |>
  filter(system_name %in% system_total$system_name) |>
  mutate(system_name = factor(system_name, levels = system_total$system_name))

# Horizontal stacked bar chart for top systems
p_system_hazard <- ggplot(
  system_analytes_top,
  aes(y = system_name,
      x = total_hazard_analyte,
      fill = analyte)
) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = hazard_colors) +
  labs(
    title = "Hazard Index Decomposition for Top Systems",
    x = "Hazard Index",
    y = "System Name",
    fill = "Analyte"
  ) +
  theme(
    legend.position = "bottom"
  )

ggsave("system_hazard.svg", p_system_hazard, width = 14, height = 10)
```

```{r}
#| label: pfas sites
# Static PFAS sites map
pfas_sites_sf <- pfas_sites %>%
  mutate(
    type = str_replace_all(type, "mlitary", "military"),
    type = str_to_title(type),
    type = str_replace_all(type, "Laundromat/Drycleaner", "Laundromat/Dry Cleaner"),
    type = case_when(
      type %in% c("Tannery", "Unknown", "Refinery", "Laundromat/Dry Cleaner", "Paper Manufacturing", "Fire Station") ~ "Other",
      TRUE ~ type
    )
  ) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

pfas_static_map2 <- ggplot() +
  geom_sf(data = mi_counties,
          fill = "white",
          color = "black",
          linewidth = 0.3) +
  geom_sf(data = pfas_sites_sf,
          aes(fill = type),
          size = 5,
          pch = 21,
          alpha = 0.7,
          color = "black") +
scale_fill_manual(values = c(
    "Landfill"     = "#E69F00",
    "Industrial"   = "#56B4E9",
    "Other"        = "#CC79A7",
    "Plating"      = "#F0E442",
    "Wastewater"   = "#0072B2",
    "Airport"      = "#999999",
    "Military"     = "#009E73",
    "Fire Related" = "#D55E00"
)) +
  labs(fill = "Site Type") +
  coord_sf(
  xlim = c(mi_bbox["xmin"], mi_bbox["xmax"]),
  ylim = c(mi_bbox["ymin"], mi_bbox["ymax"]),
  expand = FALSE
) +
  theme_map(base_size = 32) +
theme(
    legend.position = c(0.02, 0.1),
    legend.justification = c("left", "bottom"),
    legend.key.size = unit(1.2, "cm"),
    legend.spacing.x = unit(0.4, "cm"),
    plot.margin = ggplot2::margin(0, 0, 0, 0),
    text = element_text(family = "Arial"),
    legend.title = element_text(face = "bold", size = 22, color = "black"),
    legend.text  = element_text(face = "bold", size = 20, color = "black")
  )

ggsave("pfas_sites_map2.svg", pfas_static_map2, width = 15, height = 18)
```

