---
title: "PFAS Modeling"
format:
  dashboard:
    orientation: rows   
    theme: flatly          
    toc: true              
    self-contained: true   
execute:
  echo: false
  warning: false
  message: false
  output: true
---

```{r}
library(tidyverse)
library(readr)
library(janitor)
library(sf)
```

```{r}
# Load all final cleaned datasets
surface_water1 <- read_csv("surface_water1.csv")
pfas_sites1    <- read_csv("pfas_sites1.csv")
public_water1  <- read_csv("public_water1.csv")
merged_data1   <- read_csv("merged_data1.csv")
```

```{r}
# Load the aquifer polygon layer. The shapefile's .dbf carries the same OBJECTID/AQU_UNIT/AQU_CHAR fields.
aquifer <- st_read("Aquifer_Characteristics_Of_Glacial_Drift.shp", quiet = TRUE)
```

```{r}
glimpse(aquifer)
```

```{r}
glimpse(merged_data1)
```

```{r}
glimpse(surface_water1)
```

```{r}
glimpse(public_water1)
```

```{r}
glimpse(pfas_sites1)
```

## Spatial Joins of aquifer, public water and pfas sites

### Attaching aquifer characteristics to public_water1

```{r}
# Joining datasets
water_locations <- public_water1 |>
  distinct(hex_id, longitude, latitude) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) # changing the data to spatial objects

# Transforming to CRS from 4326 to aquifer
water_locations_proj <- st_transform(water_locations, st_crs(aquifer))
```

```{r}
## Spatially join aquifer attributes onto each location, then drop geometry to get a plain hex_id lookup table
water_aquifer <- water_locations_proj |>
  st_join(aquifer |> select(AQU_UNIT, AQU_CHAR)) |>
  st_drop_geometry() |>
  select(hex_id, AQU_UNIT, AQU_CHAR)
```

```{r}
# Check match rate before proceeding, the total locations joined, those that dont have aquifer match 
water_aquifer |>
  summarise(total = n(),
            missing = sum(is.na(AQU_UNIT)),
            pct_missing = round(mean(is.na(AQU_UNIT)) * 100, 1))
```

```{r}
# Pull out the locations that failed to match an aquifer polygon
unmatched <- water_locations_proj |>
  st_join(aquifer |> select(AQU_UNIT, AQU_CHAR)) |>
  filter(is.na(AQU_UNIT))

nrow(unmatched)   # should be 77

# Look at where they are
unmatched |>
  st_drop_geometry() |>
  select(hex_id, longitude, latitude)
```

```{r}
ggplot() +
  geom_sf(data = aquifer, fill = "grey90", color = "grey60") +
  geom_sf(data = unmatched, color = "#eb6834", size = 1.5) +
  labs(title = "Public-water locations with no aquifer match") +
  theme_minimal()
```

### Connect public water to nearby PFAS sites (distance-based, not geoid)

```{r}
# Convert PFAS sites to sf points, reproject to match the aquifer's projected CRS so distance calculations are in meters, not degrees
pfas_sites_sf   <- pfas_sites1 |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)
pfas_sites_proj <- st_transform(pfas_sites_sf, st_crs(aquifer))
```

```{r}
# Nearest PFAS site to each public-water location(connecting public water to pfas site)
# For every water location check across PFAS sites and return the closest one
nearest_idx  <- st_nearest_feature(water_locations_proj, pfas_sites_proj)
nearest_dist <- st_distance(water_locations_proj, pfas_sites_proj[nearest_idx, ], by_element = TRUE)

water_pfas_nearest <- water_locations_proj |>
  st_drop_geometry() |>
  select(hex_id) |>
  mutate(
    distance_to_nearest_site_m = as.numeric(nearest_dist),
    nearest_site_facility = pfas_sites_proj$facility[nearest_idx],
    nearest_site_type     = pfas_sites_proj$type[nearest_idx]
  )
```

```{r}
# Count of PFAS sites within a set radius (e.g. 5 km) -- adjust threshold as needed
within_5km <- st_is_within_distance(water_locations_proj, pfas_sites_proj, dist = 5000)

water_pfas_nearest <- water_pfas_nearest |>
  mutate(
    n_sites_within_5km = lengths(within_5km),
    pfas_site_nearby    = n_sites_within_5km > 0
  )
```

```{r}
# Classify PFAS sites themselves by aquifer, so you can ask should answer questions like "are known sites concentrated in certain aquifer types?"
pfas_sites_aquifer <- pfas_sites_proj |>
  st_join(aquifer |> select(AQU_UNIT, AQU_CHAR)) |>
  st_drop_geometry()
```

### Combine into one joined dataset

```{r}
# One row per public-water observation per hex_id x analyte with aquifer characteristics and proximity-based with known PFAS sites
modeling_data <- public_water1 |>
  left_join(water_aquifer, by = "hex_id") |>
  left_join(water_pfas_nearest, by = "hex_id")

glimpse(modeling_data)
```

```{r}
write_csv(modeling_data, "modeling_data.csv")
```

Reencoding the Aquifer Characteristics

```{r}

```

## Exploratory Analysis of the Joined Dataset

```{r}
# Create location level object
location_level <- modeling_data |>
  distinct(hex_id, system_name, system_type, Hazard_Index,
           AQU_UNIT, AQU_CHAR, distance_to_nearest_site_m,
           nearest_site_type, n_sites_within_5km, pfas_site_nearby)

# Confirm there is "one value per location" 
location_level |>
  count(hex_id) |>
  filter(n > 1)

nrow(location_level)
```

### Aquifer coverage

```{r}
location_level |>
  summarise(
    total_locations = n(),
    missing_aquifer  = sum(is.na(AQU_UNIT)),
    pct_missing      = round(mean(is.na(AQU_UNIT)) * 100, 1)
  )
```

```{r}
location_level |>
  count(AQU_CHAR, sort = TRUE)
```

### Hazard Index by aquifer type

```{r}
hazard_by_aquifer <- location_level |>
  filter(!is.na(AQU_CHAR)) |>
  group_by(AQU_CHAR) |>
  summarise(
    mean_hazard = mean(Hazard_Index, na.rm = TRUE),
    n_locations = n(),
    .groups = "drop"
  ) |>
  arrange(desc(mean_hazard))

hazard_by_aquifer
```

```{r}
ggplot(hazard_by_aquifer, aes(x = reorder(AQU_CHAR, mean_hazard), y = mean_hazard)) +
  geom_col(fill = "#2a78d6", width = 0.7) +
  coord_flip() +
  labs(
    title = "Mean Hazard Index by Aquifer Characteristic",
    x = NULL, y = "Mean Hazard Index"
  ) +
  theme_minimal()
```

### Proximity to known PFAS sites

```{r}
location_level |>
  group_by(pfas_site_nearby) |>
  summarise(
    mean_hazard   = mean(Hazard_Index, na.rm = TRUE),
    median_hazard = median(Hazard_Index, na.rm = TRUE),
    n_locations   = n(),
    .groups = "drop"
  )
```

```{r}
ggplot(location_level, aes(x = pfas_site_nearby, y = Hazard_Index, fill = pfas_site_nearby)) +
  geom_boxplot(width = 0.5) +
  scale_fill_manual(values = c("TRUE" = "#eb6834", "FALSE" = "#2a78d6"), guide = "none") +
  labs(
    title = "Hazard Index: Locations Near vs. Not Near a Known PFAS Site",
    x = "PFAS site within 5 km", y = "Hazard Index"
  ) +
  theme_minimal()
```

```{r}
ggplot(location_level, aes(x = distance_to_nearest_site_m, y = Hazard_Index)) +
  geom_point(color = "#2a78d6", alpha = 0.5) +
  labs(
    title = "Hazard Index vs. Distance to Nearest Known PFAS Site",
    x = "Distance to nearest site (m)", y = "Hazard Index"
  ) +
  theme_minimal()
```

### Aquifer type x PFAS-site proximity

```{r}
location_level |>
  filter(!is.na(AQU_CHAR)) |>
  count(AQU_CHAR, pfas_site_nearby) |>
  pivot_wider(names_from = pfas_site_nearby, values_from = n, values_fill = 0)
```