---
title: "PFAS Modeling"
format:
  dashboard:
    orientation: rows   
    theme: flatly          
    toc: true              
    self-contained: true   
execute:
  echo: false
  warning: false
  message: false
  output: true
---

```{r}
library(tidyverse)
library(readr)
library(janitor)
library(sf)
library(tidymodels)
```

```{r}
# Load the cleaned source datasets
surface_water1 <- read_csv("surface_water1.csv")
pfas_sites1    <- read_csv("pfas_sites1.csv")
public_water1  <- read_csv("public_water1.csv")
merged_data1   <- read_csv("merged_data1.csv")
```

```{r}
# Load the aquifer polygon layer (OBJECTID/AQU_UNIT/AQU_CHAR fields)
aquifer <- st_read("Aquifer_Characteristics_Of_Glacial_Drift.shp", quiet = TRUE)
```

```{r}
glimpse(aquifer)
```

```{r}
glimpse(merged_data1)
```

```{r}
glimpse(surface_water1)
```

```{r}
glimpse(public_water1)
```

```{r}
glimpse(pfas_sites1)
```

## Spatial Joins of aquifer, public water and pfas sites

### Attaching aquifer characteristics to public_water1

```{r}
# Collapse to one row per unique sampling location and convert to spatial points
water_locations <- public_water1 |>
  distinct(hex_id, longitude, latitude) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)

water_locations_proj <- st_transform(water_locations, st_crs(aquifer))
```

```{r}
# Attach aquifer attributes to each location by spatial join, then drop back to a plain table
water_aquifer <- water_locations_proj |>
  st_join(aquifer |> select(AQU_UNIT, AQU_CHAR)) |>
  st_drop_geometry() |>
  select(hex_id, AQU_UNIT, AQU_CHAR)
```

```{r}
# n_locations = total sampling locations; n_missing_aquifer = locations with no aquifer match
water_aquifer |>
  summarise(
    n_locations       = n(),
    n_missing_aquifer = sum(is.na(AQU_UNIT)),
    pct_missing       = round(mean(is.na(AQU_UNIT)) * 100, 1)
  )
```

```{r}
# Pull out the locations that failed to match an aquifer polygon
unmatched <- water_locations_proj |>
  st_join(aquifer |> select(AQU_UNIT, AQU_CHAR)) |>
  filter(is.na(AQU_UNIT))

nrow(unmatched)   # should be 77

unmatched |>
  st_drop_geometry() |>
  select(hex_id, longitude, latitude)
```

```{r}
# Map the unmatched points to confirm they sit on Great Lakes shorelines, not a join bug
ggplot() +
  geom_sf(data = aquifer, fill = "grey90", color = "grey60") +
  geom_sf(data = unmatched, color = "#eb6834", size = 1.5) +
  labs(title = "Public-water locations with no aquifer match") +
  theme_minimal()
```

### Connect public water to nearby PFAS sites (distance-based metric)

```{r}
# Convert PFAS sites to spatial points, reproject to meters to match the aquifer CRS
pfas_sites_sf   <- pfas_sites1 |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)
pfas_sites_proj <- st_transform(pfas_sites_sf, st_crs(aquifer))
```

```{r}
# For every water location, find the single closest PFAS site and its distance in meters
nearest_idx  <- st_nearest_feature(water_locations_proj, pfas_sites_proj)
nearest_dist <- st_distance(water_locations_proj, pfas_sites_proj[nearest_idx, ], by_element = TRUE)

water_pfas_nearest <- water_locations_proj |>
  st_drop_geometry() |>
  select(hex_id) |>
  mutate(
    distance_to_nearest_site_m = as.numeric(nearest_dist),
    nearest_site_facility = pfas_sites_proj$facility[nearest_idx],
    nearest_site_type     = pfas_sites_proj$type[nearest_idx]
  )
```

```{r}
# Count how many PFAS sites (not just the nearest one) fall within a 5km radius
within_5km <- st_is_within_distance(water_locations_proj, pfas_sites_proj, dist = 5000)

water_pfas_nearest <- water_pfas_nearest |>
  mutate(
    n_sites_within_5km = lengths(within_5km),
    pfas_site_nearby    = n_sites_within_5km > 0
  )
```

```{r}
# Classify the PFAS sites themselves by aquifer, for a sites-level (not water-level) view
pfas_sites_aquifer <- pfas_sites_proj |>
  st_join(aquifer |> select(AQU_UNIT, AQU_CHAR)) |>
  st_drop_geometry()
```

### Combine into one joined dataset

```{r}
# Attach the aquifer and PFAS-proximity lookup tables onto every public_water1 row
modeling_data <- public_water1 |>
  left_join(water_aquifer, by = "hex_id") |>
  left_join(water_pfas_nearest, by = "hex_id")

glimpse(modeling_data)
```

```{r}
write_csv(modeling_data, "modeling_data.csv")
```

Reencoding the Aquifer Characteristics

```{r}
# Inspect the raw aquifer codes and their full text descriptions before recoding
modeling_data |>
  distinct(AQU_UNIT, AQU_CHAR) |>
  arrange(AQU_UNIT)
```

```{r}
# Map each AQU_UNIT code to a short, chart-friendly ordered label
modeling_data <- modeling_data |>
  mutate(
    aquifer_label = case_when(
      AQU_UNIT == "1"               ~ "I: Thin drift over bedrock",
      AQU_UNIT == "2"               ~ "II: Drift generally not an aquifer",
      AQU_UNIT %in% c("3a", "3aX")  ~ "IIIA: Interbedded, unconfined near surface",
      AQU_UNIT %in% c("3b", "3bX")  ~ "IIIB: Interbedded, variable near surface",
      AQU_UNIT == "4"               ~ "IV: Possible aquifer, undelineated",
      TRUE ~ NA_character_
    ),
    aquifer_label = factor(
      aquifer_label,
      levels = c(
        "I: Thin drift over bedrock",
        "II: Drift generally not an aquifer",
        "IIIA: Interbedded, unconfined near surface",
        "IIIB: Interbedded, variable near surface",
        "IV: Possible aquifer, undelineated"
      )
    )
  )

# n_locations = distinct hex_id sampling locations in each aquifer label
modeling_data |>
  distinct(hex_id, aquifer_label) |>
  count(aquifer_label, name = "n_locations", sort = TRUE)
```

```{r}
# Apply the same recoding to the PFAS-sites table (separate table, same lookup logic)
pfas_sites_aquifer <- pfas_sites_aquifer |>
  mutate(
    aquifer_label = case_when(
      AQU_UNIT == "1"               ~ "I: Thin drift over bedrock",
      AQU_UNIT == "2"               ~ "II: Drift generally not an aquifer",
      AQU_UNIT %in% c("3a", "3aX")  ~ "IIIA: Interbedded, unconfined near surface",
      AQU_UNIT %in% c("3b", "3bX")  ~ "IIIB: Interbedded, variable near surface",
      AQU_UNIT == "4"               ~ "IV: Possible aquifer, undelineated",
      TRUE ~ NA_character_
    )
  )

# n_sites = known PFAS sites in each aquifer label
pfas_sites_aquifer |>
  count(aquifer_label, name = "n_sites") |>
  arrange(aquifer_label)
```

## Exploratory Analysis of the Joined Dataset

Everything in this section works at the **hex_id which is the sampling location** for location-level modeling, before the system level analysis of the exact public water drinking sites are analysed

```{r}
# Collapse modeling_data down to one row per sampling location (drop the analyte-level rows)
location_level <- modeling_data |>
  distinct(hex_id, system_name, system_type, system_short, Hazard_Index,
           AQU_UNIT, aquifer_label, distance_to_nearest_site_m,
           nearest_site_type, n_sites_within_5km, pfas_site_nearby)

# QA: n_rows_per_hex should be 1 for every hex_id (confirms one row per location)
location_level |>
  count(hex_id, name = "n_rows_per_hex") |>
  filter(n_rows_per_hex > 1)

nrow(location_level)
```

This chunk collapses modeling_data from 45,080 analyte-level rows down to 1,610 rows — one per sampling location by keeping only the columns that don't vary by analyte (Hazard_Index, aquifer, distance, etc.) and dropping the duplicate rows that resulted from testing 28 different chemicals at each location.No hex_id ended up with more than one row and the final count is 1,610 rows.

### Understanding the location hierarchy: hex_id -> system_name -> system_type

```{r}
# hex_id = one sampling spot; system_name = the utility it belongs to; system_type = its category
location_level |>
  summarise(
    n_hex_ids       = n_distinct(hex_id),
    n_water_systems = n_distinct(system_name),
    n_system_types  = n_distinct(system_type)
  )
```

There are water locations that have more than one sampling location that is why the hex-ids dont match the sytem name

```{r}
# Each hex_id should have exactly one system_name, each system_name exactly one system_type
location_level |>
  group_by(hex_id) |>
  summarise(n_system_names = n_distinct(system_name), .groups = "drop") |>
  filter(n_system_names > 1)   
location_level |>
  group_by(system_name) |>
  summarise(n_system_types = n_distinct(system_type), .groups = "drop") |>
  filter(n_system_types > 1)   
```

Confirms that every system name or water sytem has one sytem type

```{r}
# n_hex_ids = sampling locations per water system; summarised to see the typical spread
location_level |>
  count(system_name, name = "n_hex_ids") |>
  summarise(
    min_locations    = min(n_hex_ids),
    median_locations = median(n_hex_ids),
    max_locations    = max(n_hex_ids)
  )
```

Most water syetms have 1 sampling location however there are some that have upto 9 locations that is why we shrink it to the highest 

### Aquifer coverage

```{r}
# n_locations = all sampling locations; n_missing_aquifer = those with no aquifer_label
location_level |>
  summarise(
    n_locations       = n(),
    n_missing_aquifer = sum(is.na(aquifer_label)),
    pct_missing       = round(mean(is.na(aquifer_label)) * 100, 1)
  )
```

```{r}
# n_locations = distinct sampling locations per aquifer type
location_level |>
  count(aquifer_label, name = "n_locations") |>
  arrange(aquifer_label)
```

### Hazard Index by aquifer type 

```{r}
### Hazard Index by aquifer type

# pct_detected = share of locations with ANY detectable PFAS (median is uninformative
# here since 86% of all locations read exactly 0, so median = 0 for every aquifer type)
# p90_hazard = 90th percentile, a "how bad for the worse 10%" measure median can't give
hazard_by_aquifer <- location_level |>
  filter(!is.na(aquifer_label)) |>
  group_by(aquifer_label) |>
  summarise(
    pct_detected  = round(100 * mean(Hazard_Index > 0, na.rm = TRUE), 1),
    mean_hazard   = round(mean(Hazard_Index, na.rm = TRUE), 2),
    p90_hazard    = round(quantile(Hazard_Index, 0.90, na.rm = TRUE), 3),
    max_value     = round(max(Hazard_Index, na.rm = TRUE), 2),
    n_locations   = n(),
    water_systems = n_distinct(system_name),
    .groups = "drop"
  ) |>
  arrange(aquifer_label)

hazard_by_aquifer
```

```{r}
# Mean Hazard Index per aquifer type, sensitive to a few extreme hotspot locations
ggplot(hazard_by_aquifer, aes(x = aquifer_label, y = mean_hazard)) +
  geom_col(fill = "darkorange", width = 0.7) +
  coord_flip() +
  scale_x_discrete(limits = rev) +
  labs(title = "Mean Hazard Index by Aquifer Type", x = NULL, y = "Mean Hazard Index") +
  theme_minimal()
```

```{r}
# Share of locations with any detectable PFAS -- replaces median, which is 0 everywhere
ggplot(hazard_by_aquifer, aes(x = aquifer_label, y = pct_detected)) +
  geom_col(fill = "#2a78d6", width = 0.7) +
  coord_flip() +
  scale_x_discrete(limits = rev) +
  labs(title = "Percent of Locations with Detectable PFAS by Aquifer Type", x = NULL, y = "% Locations Detected") +
  theme_minimal()
```

```{r}
# 90th percentile Hazard Index per aquifer type -- how bad it gets for the worse 10%
ggplot(hazard_by_aquifer, aes(x = aquifer_label, y = p90_hazard)) +
  geom_col(fill = "#1baf7a", width = 0.7) +
  coord_flip() +
  scale_x_discrete(limits = rev) +
  labs(title = "90th Percentile Hazard Index by Aquifer Type", x = NULL, y = "90th Percentile Hazard Index") +
  theme_minimal()
```

```{r}
# Max Hazard Index per aquifer type, the worst single location, a precautionary read
ggplot(hazard_by_aquifer, aes(x = aquifer_label, y = max_value)) +
  geom_col(fill = "#eb6834", width = 0.7) +
  coord_flip() +
  scale_x_discrete(limits = rev) +
  labs(title = "Max Hazard Index by Aquifer Type", x = NULL, y = "Max Hazard Index") +
  theme_minimal()
```

mean and pct_detected tell you how widespread contamination is within each aquifer type, while p90 and max tell you how severe it gets at the tail. Reading them together: an aquifer type with low pct_detected but a high max_value means it's mostly clean with one bad outlier; high pct_detected and high p90 together means contamination is broadly present, not just a fluke.

### Hazard Index by system type 

```{r}
# system_short = short code (e.g. SCH, MUN) standing in for the long system_type description
hazard_by_system_type <- location_level |>
  group_by(system_short) |>
  summarise(
    mean_hazard   = round(mean(Hazard_Index, na.rm = TRUE), 2),
    median_hazard = round(median(Hazard_Index, na.rm = TRUE), 2),
    n_locations   = n(),
    water_systems = n_distinct(system_name),
    .groups = "drop"
  ) |>
  arrange(desc(mean_hazard))

hazard_by_system_type
```

```{r}
ggplot(hazard_by_system_type, aes(x = reorder(system_short, mean_hazard), y = mean_hazard)) +
  geom_col(fill = "darkorange", width = 0.7) +
  coord_flip() +
  labs(
    title = "Mean Hazard Index by System Type",
    x = NULL, y = "Mean Hazard Index"
  ) +
  theme_minimal()
```

### What the PFAS-site distance metrics actually measure

```{r}
# distance_to_nearest_site_m = straight-line distance to the single closest known PFAS site
location_level |>
  summarise(
    min_dist_m    = round(min(distance_to_nearest_site_m, na.rm = TRUE), 0),
    median_dist_m = round(median(distance_to_nearest_site_m, na.rm = TRUE), 0),
    mean_dist_m   = round(mean(distance_to_nearest_site_m, na.rm = TRUE), 0),
    max_dist_m    = round(max(distance_to_nearest_site_m, na.rm = TRUE), 0)
  )
```

```{r}
# n_locations = sampling locations with vs. without a known site within 5km
location_level |>
  count(pfas_site_nearby, name = "n_locations") |>
  mutate(pct = round(100 * n_locations / sum(n_locations), 1))
```

```{r}
# n_locations = sampling locations whose single nearest site is of that type
location_level |>
  filter(!is.na(nearest_site_type)) |>
  count(nearest_site_type, name = "n_locations", sort = TRUE)
```

### Proximity to known PFAS sites (location grain)

```{r}
# water_systems = distinct utilities with vs. without a known site nearby
location_level |>
  group_by(pfas_site_nearby) |>
  summarise(
    mean_hazard   = mean(Hazard_Index, na.rm = TRUE),
    median_hazard = median(Hazard_Index, na.rm = TRUE),
    n_locations   = n(),
    water_systems = n_distinct(system_name),
    .groups = "drop"
  )
```

```{r}
ggplot(location_level, aes(x = pfas_site_nearby, y = Hazard_Index, fill = pfas_site_nearby)) +
  geom_boxplot(width = 0.5) +
  scale_fill_manual(values = c("TRUE" = "#eb6834", "FALSE" = "#2a78d6"), guide = "none") +
  labs(
    title = "Hazard Index: Locations Near vs. Not Near a Known PFAS Site",
    x = "PFAS site within 5 km", y = "Hazard Index"
  ) +
  theme_minimal()
```

```{r}
ggplot(location_level, aes(x = distance_to_nearest_site_m, y = Hazard_Index)) +
  geom_point(color = "#2a78d6", alpha = 0.5) +
  labs(
    title = "Hazard Index vs. Distance to Nearest Known PFAS Site",
    x = "Distance to nearest site (m)", y = "Hazard Index"
  ) +
  theme_minimal()
```

### Aquifer type x PFAS-site proximity (location grain)

```{r}
# n_locations = sampling locations in each aquifer x proximity combination
location_level |>
  filter(!is.na(aquifer_label)) |>
  count(aquifer_label, pfas_site_nearby, name = "n_locations") |>
  pivot_wider(names_from = pfas_site_nearby, values_from = n_locations, values_fill = 0)
```

```{r}
# Same cross-tab, but water_systems = distinct utilities instead of sampling locations
location_level |>
  filter(!is.na(aquifer_label)) |>
  distinct(system_name, aquifer_label, pfas_site_nearby) |>
  count(aquifer_label, pfas_site_nearby, name = "water_systems") |>
  pivot_wider(names_from = pfas_site_nearby, values_from = water_systems, values_fill = 0)
```

## Water-System Level Analysis

The project's real question is how badly **public water systems** This section rolls `location_level` up to
one row per `system_name`, using `max_hazard` (the system's worst sampling point) as
the headline number, since a public-health read should not let a few clean samples
average away a real hotspot.

### Rolling locations up to systems

```{r}
# One row per system_name: worst reading, typical reading, spread, and closest known-site distance
system_hazard <- location_level |>
  group_by(system_name, system_type) |>
  summarise(
    n_hex_ids               = n(),
    max_hazard              = round(max(Hazard_Index, na.rm = TRUE), 2),
    mean_hazard             = round(mean(Hazard_Index, na.rm = TRUE), 2),
    hazard_range            = round(max_hazard - min(Hazard_Index, na.rm = TRUE), 2),
    min_distance_to_site_m  = round(min(distance_to_nearest_site_m, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  arrange(desc(max_hazard))

nrow(system_hazard)   # should equal n_water_systems, not n_hex_ids
system_hazard
```

```{r}
# Top 15 worst-affected systems by their single highest Hazard Index reading
top_systems <- system_hazard |> slice_head(n = 15)

ggplot(top_systems, aes(x = reorder(system_name, max_hazard), y = max_hazard)) +
  geom_col(fill = "#eb6834") +
  coord_flip() +
  labs(
    title = "Water Systems with the Highest Hazard Index",
    x = NULL, y = "Max Hazard Index (worst sampling location)"
  ) +
  theme_minimal()
```

```{r}
# n_systems_above_alert = systems with at least one reading at/above the EPA limit
# (Hazard_Index = 1 means combined exposure across the four PFAS chemicals equals
# their EPA health-based reference level; above 1 means it exceeds that limit)
HAZARD_ALERT_THRESHOLD <- 1

system_hazard |>
  summarise(
    n_water_systems       = n(),
    n_systems_above_alert = sum(max_hazard > HAZARD_ALERT_THRESHOLD),
    pct_above_alert       = round(100 * mean(max_hazard > HAZARD_ALERT_THRESHOLD), 1)
  )
```

### Does spread between a system's own locations relate to its Hazard Index variability?

```{r}
# Only systems with more than one sampling location have a "spread" to measure
multi_location_systems <- location_level |>
  group_by(system_name) |>
  filter(n() > 1) |>
  summarise(
    n_hex_ids    = n(),
    hazard_range = round(max(Hazard_Index) - min(Hazard_Index), 2),
    hazard_sd    = round(sd(Hazard_Index), 2),
    .groups = "drop"
  )

nrow(multi_location_systems)   # how many systems this check even applies to
```

```{r}
# max_spread_m = farthest distance between any two of a system's own sampling points
system_spread <- water_locations_proj |>
  left_join(public_water1 |> distinct(hex_id, system_name), by = "hex_id") |>
  semi_join(multi_location_systems, by = "system_name") |>
  group_by(system_name) |>
  group_modify(~ tibble(max_spread_m = as.numeric(max(st_distance(.x))))) |>
  ungroup()

multi_location_systems |>
  left_join(system_spread, by = "system_name") |>
  arrange(desc(hazard_range))
```

## Data Preparation for Modeling 

### Outcome variable: log(max Hazard Index)

```{r}
# log1p handles Hazard_Index == 0 safely (log(0) is undefined, log1p(0) = 0)
model_data <- system_hazard |>
  filter(!is.na(min_distance_to_site_m)) |>
  mutate(log_hazard = log1p(max_hazard))
```

```{r}
# Compare raw vs. log-transformed distributions to check the log fixed the right-skew
model_data |>
  summarise(
    mean_hazard        = round(mean(max_hazard), 2),
    median_hazard      = round(median(max_hazard), 2),
    mean_log_hazard    = round(mean(log_hazard), 2),
    median_log_hazard  = round(median(log_hazard), 2)
  )
```

```{r}
ggplot(model_data, aes(x = log_hazard)) +
  geom_histogram(fill = "#2a78d6", bins = 30) +
  labs(
    title = "Distribution of log(1 + Max Hazard Index) by Water System",
    x = "log(1 + Max Hazard Index)", y = "Number of water systems"
  ) +
  theme_minimal()
```

### Choosing the distance cutoff empirically

```{r}
# Full distribution of each system's closest-known-site distance, to justify the cutoff
model_data |>
  summarise(
    p25    = round(quantile(min_distance_to_site_m, 0.25, na.rm = TRUE)),
    median = round(quantile(min_distance_to_site_m, 0.50, na.rm = TRUE)),
    p75    = round(quantile(min_distance_to_site_m, 0.75, na.rm = TRUE)),
    p90    = round(quantile(min_distance_to_site_m, 0.90, na.rm = TRUE))
  )
```

```{r}
# pct_capped = share of systems that would be pulled down to each candidate cutoff
candidate_cutoffs <- tibble(cutoff_km = c(5, 7, 10, 12, 15)) |>
  mutate(cutoff_m = cutoff_km * 1000)

candidate_cutoffs$pct_capped <- sapply(candidate_cutoffs$cutoff_m, function(c) {
  round(100 * mean(model_data$min_distance_to_site_m > c, na.rm = TRUE), 1)
})

candidate_cutoffs
```

```{r}
# 10km sits at the median ~50th percentile of systems' closest-site distance, splitting them roughly evenly into "closer than typical" vs "farther than typical" 
DISTANCE_CUTOFF_M <- 10000

model_data <- model_data |>
  mutate(distance_capped_m = pmin(min_distance_to_site_m, DISTANCE_CUTOFF_M))
```

```{r}
# n_capped = how many systems actually got pulled down to the cutoff
model_data |>
  summarise(
    n_water_systems = n(),
    n_capped        = sum(min_distance_to_site_m > DISTANCE_CUTOFF_M),
    pct_capped      = round(100 * mean(min_distance_to_site_m > DISTANCE_CUTOFF_M), 1)
  )
```

```{r}
ggplot(model_data, aes(x = min_distance_to_site_m, y = distance_capped_m)) +
  geom_point(color = "#2a78d6", alpha = 0.4) +
  geom_hline(yintercept = DISTANCE_CUTOFF_M, color = "#eb6834", linetype = "dashed") +
  labs(
    title = "Raw vs. Tapered (Capped) Distance to Nearest PFAS Site",
    x = "Raw distance (m)", y = "Capped distance (m)"
  ) +
  theme_minimal()
```

### Simple linear regression: log(Hazard Index) on tapered distance 

```{r}
# Recipe: formalizes the same log-transform and distance-cap used above, inside the modeling pipeline
pfas_recipe <- recipe(max_hazard ~ min_distance_to_site_m, data = model_data) |>
  step_mutate(distance_capped_m = pmin(min_distance_to_site_m, DISTANCE_CUTOFF_M)) |>
  step_rm(min_distance_to_site_m) |>
  step_log(max_hazard, offset = 1)
```

```{r}
# Model spec: ordinary least squares via the "lm" engine, one predictor = "simple" regression
lm_spec <- linear_reg() |> set_engine("lm")
```

```{r}
# Workflow bundles the recipe (preprocessing) and the model spec together
pfas_workflow <- workflow() |>
  add_recipe(pfas_recipe) |>
  add_model(lm_spec)

pfas_fit <- fit(pfas_workflow, data = model_data)
```

```{r}
# tidy() = coefficient table (term, estimate, std.error, statistic, p.value)
tidy(pfas_fit)
```

```{r}
# glance() = one-row model fit summary (R-squared, adjusted R-squared, etc.)
glance(pfas_fit)
```

```{r}
ggplot(model_data, aes(x = distance_capped_m, y = log_hazard)) +
  geom_point(color = "#2a78d6", alpha = 0.4) +
  geom_smooth(method = "lm", color = "#eb6834", se = TRUE) +
  labs(
    title = "log(Max Hazard Index) vs. Tapered Distance to Nearest PFAS Site",
    x = "Capped distance to nearest site (m)", y = "log(1 + Max Hazard Index)"
  ) +
  theme_minimal()
```